---
title: "Projeto1"
author: "Lucas Miranda de Sena"
date: '2023-01-25'
output: html_document
---
EXPLORATORY DATA ANALYSIS

PACKAGES
```{r, echo=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggridges)
library(gridExtra)
```

READING DATA
```{r}
data <- read.csv("heart.csv",
                           header = T,
                           sep = ",",
                           dec = ".")
```

DATASET INFO
```{r}
data %>% str()
```
The dataset contains informations of 303 pacients and 13 features about their heart system conditions. There is also one last feature with the diagnosis of a possible heart disease. 

Dataset description

Descrição das variáveis.
age: Age of the patient
sex: Sex of the patient
cp: Chest Pain type
   0: typical angina
   1: atypical angina
   2: non-anginal pain
   3: asymptomatic
trtbps: resting blood pressure (mmHg)
chol: cholesterol (mg/dl)
fbs: fasting blood sugar > 120 mg/dl
   1 = true
   0 = false
restecg: resting electrocardiographic results   
   0: normal
   1: having ST-T wave abnormality (T wave inversions and/or ST elevation or depression of > 0.05 mV)
   2: showing probable or definite left ventricular hypertrophy by Estes' criteria
thalachh: maximum heart rate achieved
exng: exercise induced angina
   1 = yes
   0 = no
oldpeak: ST depression induced by exercise relative to rest
slp: the slope of the peak exercise ST segment
   0 = unsloping
   1 = flat
   2 = downsloping
caa: number of major vessels (0-3)
thall: thalassemia
   0 = null
   1 = fixed defect
   2 = normal
   3 = reversable defect
output: diagnosis of heart disease
   0: < 50% diameter narrowing. less chance of heart disease
   1: > 50% diameter narrowing. more chance of heart disease

   
There are no missing values in this dataset and all the features are numerical (integers and floats). 
However, some features are integers, but the numbers represents categories. In order to peform a good EDA, a data transformation is necessary.


DATA TRANSFORMATION
```{r}
data_eda <- data
```

```{r}
data_eda$sex <- data_eda$sex %>% 
  factor(levels = c(1, 0),
         labels = c('M', 'F'))

data_eda <- data_eda %>% 
  rename('chest_pain_type' = 'cp')
data_eda$chest_pain_type <- data_eda$chest_pain_type %>% 
  factor(levels = c(0, 1, 2, 3),labels = c('typical angina',
                                           'atypical angina',
                                           'non-anginal pain',
                                           'asymptomatic'))

data_eda <- data_eda %>% 
  rename('fasting_blood_sugar' = 'fbs')
data_eda$fasting_blood_sugar <- data_eda$fasting_blood_sugar %>% 
  factor(levels = c(1, 0),
         labels = c('yes',
                    'no'))

data_eda$restecg <- data_eda$restecg %>% 
  factor(levels = c(0,1,2),
         labels = c('normal',
                    'wave abnormality',
                    'probable left ventricular hypertrophy'))

data_eda$exng <- data_eda$exng %>% 
  factor(levels = c(0,1),
         labels = c('yes',
                    'no'))

data_eda$slp <- data_eda$slp %>% 
  factor(levels = c(0,1,2),
         labels = c('usloping',
                    'flat',
                    'downsloping'))

data_eda$thall <- data_eda$thall %>% 
  factor(levels = c(0,1,2,3),
         labels = c('null',
                    'fixed defect',
                    'normal', 
                    'reversable defect'))

data_eda$output <- data_eda$output %>%
  factor(levels = c(0,1),
         labels = c('Less chance of disease',
                    'More chance of disease'))
```

RENAMING FEATURES
```{r}
data_eda <- data_eda %>%
  rename('cholesterol' = 'chol',
         'max_heart_rate' = 'thalachh',
         'slope' = 'slp',
         'number_of_major_vessels' = 'caa',
         'diagnosis' = 'output')
```

```{r}
data_eda %>% summary()
```

DATA VISUALIZATION

Exploring the correlation of age with other features
```{r}
data_eda %>%
  ggplot() +
  geom_point(aes(x = cholesterol,
                 y = age,
                 color = sex)) +
  labs(title = 'Correlation between age and cholesterol',
       x = 'Cholesterol (mg/l)',
       y = 'Age') + 
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))
```
Taking a look a the graph, it can't be cofirmed that women or men would have a higher cholesterol level.
We can take a better look at the outlier.
```{r}
data_eda %>%
  filter(cholesterol>500) %>% 
  kable(align = 'c') %>%
  kable_classic(html_font = 'courier new') %>% 
  kable_styling(bootstrap_options = 'striped',
                font_size = 14)
```


```{r}
data_eda %>% 
  ggplot() +
  geom_point(aes(y = age,
                 x = trtbps,
                 color = fasting_blood_sugar)) +
  geom_smooth(aes(y = age,
                  x= trtbps),
              method = 'lm',
              se = F) +
  labs(title = 'Correlation between age and resting blood pressure',
       x = 'Resting Blood Pressure (mmHG)',
       y = 'Age') + 
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))
```
```{r}
data_eda %>% 
  ggplot() +
  geom_point(aes(y = age,
                 x = max_heart_rate)) +
  geom_smooth(aes(y = age,
                  x = max_heart_rate),
              method = 'lm',
              se = F) +
    labs(title = 'Correlation between age and max heart rate',
       x = 'Max Heart Rate (bpm)',
       y = 'Age') +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))
```
This time, a negative correlation is seen. Elder people achive lower heart rate.

Checking the distribution of some categorical features
```{r}
data_eda %>% 
  ggplot(aes(x=age,
             y=diagnosis,
             fill = diagnosis)) +
  geom_density_ridges() +
  labs(title = 'Concentration of heart disease chance per age',
       x = 'Age',
       y = 'Diagnosis') +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))
```
A high concentration of disease chance is between 50 and 60 years old.

```{r}
data_eda %>% 
  ggplot(aes(x=age,
             y=chest_pain_type,
             fill = chest_pain_type)) +
  geom_density_ridges() +
  labs(title = 'Concentration of chest pain type per age',
       x = 'Age',
       y = 'Chest pain type') +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))
```
In the same way, there is a biger concentration in elder people.




CLUSTERING

As the symptoms shown by the pacients lead to the diagnosis of a heart disease, they can be used in a non supervised learning to cluster the patients by the simptoms. The result might be used by medical teams to choose the best treatment.

PACKAGES
```{r, echo=FALSE}
library(factoextra)
```

DATA PREPARATION
The clustering dataset wont have the diagnosis feature, and the categorical features will not be used as well.
```{r}
data_cluster <- data_eda %>% select(1,4:5,8,10,12)
data_cluster %>% head()
```

```{r}
data_cluster_scale <- data_cluster %>% scale() %>% data.frame()
```

As there are 2 diagnosis possibilities, a kmeans algorithm with 2 groups is going to be used
```{r}
set.seed(0)
cluster <- kmeans(data_cluster_scale,
                           centers = 2)
```

Result
```{r}
fviz_cluster(object = cluster,
             data = data_cluster_scale,
             repel = T,
             geom = "point") +
  labs(title = "Patient Clustering") +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect(fill = "grey97"))
```
Joining the result with the dataset
```{r}
groups <- cluster$cluster
data_comparison <- data_eda
data_comparison$cluster <- groups
data_comparison %>% head()
```

```{r}
data_comparison[,14:15] %>% table()
```
The kmeans clustering precision was 73.93%, compared to the diagnosis.


Visualizing the clustering result
```{r}

data_comparison$cluster <- data_comparison$cluster %>% factor(levels = c(1, 2),
                                                              labels = c("1", "2"))
g1 <- data_comparison %>% 
  ggplot() +
  geom_point(aes(y = age,
                 x = max_heart_rate,
                 color = diagnosis)) +
    labs(x = 'Max Heart Rate (bpm)',
         y = 'Age') +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))

g2 <- data_comparison %>% 
  ggplot() +
  geom_point(aes(y = age,
                 x = max_heart_rate,
                 color = cluster)) +
    labs(x = 'Max Heart Rate (bpm)',
         y = 'Age') +
  theme_classic(base_size = 12) +
  theme(panel.background = element_rect('gray97'))

grid.arrange(g1, g2)
```

CLASSIFICATION

PACKAGES
```{r}
library(class)
library(gmodels)
```

DATA PREPARATION
```{r}
data_classification <- data
data_classification$output <- data_classification$output %>% factor(levels = c(0,1),
                                                                labels = c('less chance of disease', 'more chance of disease'))

data_classification_scale <- data_classification[-14] %>% scale() %>% data.frame()

```

CREATING TRAINING AND TEST DATASETS
```{r}
set.seed(1)
ind <- sample(2, nrow(data_classification_scale), replace = T, prob = c(0.8, 0.2))
train <- data_classification_scale[ind == 1,]
test <- data_classification_scale[ind == 2,]

train_labels <- data_classification[ind == 1, 14]
test_labels <- data_classification[ind == 2, 14]
```

TRAINING THE MODEL AND CHECKING THE ACCURACY
```{r}
model <- knn(train = train,
             test = test,
             cl = train_labels,
             k = 15)

CrossTable(x = test_labels,
           y = model,
           prop.chisq = F)
```

The model's accuracy is 75.51%.





